Global phase diagram and possible quantum spin liquid in the triangular $J$-$K$-$\Gamma$ model

### 2020-11-11 paper/paper.tex
Paper 2020-08-11 16:10
Paper 2020-08-11 19:41

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{fig/ModelDefinition.pdf}
    \caption{\label{fig:ModelDefinition}(Color online) (a) Top view of the triangular lattice of the edge-sharing octahedron. (b) The orientation of the cubic $x$, $y$, $z$ axes with respect to the octahedron. The spin operators $S^x$, $S^y$ and $S^z$ are defined with respect to this reference frame. (c) Three types of the NN bonds on the triangular lattice, namely $\gamma=x, y, z$ colored red, green and blue, respectively. The different colors of the lattice sites label the four sublattices realizing the four sublattice transformation ( see main text for details).}
\end{figure}
In fact, magnetic ions located at the center of edge-sharing octahedra can not only form the honeycomb lattice but also the triangular lattice (see Fig.~\ref{fig:ModelDefinition}(a)), so the Kitaev and $\Gamma$ terms can naturally be generalized to the triangular lattice~\cite{PhysRevB.89.014414,PhysRevB.93.104417}. On the experimental side, studies of several classes of rare-earth compounds have recently shown that the quantum spin model on the triangular lattice can be formed by $4f$ moments. In particular, due to their possible Kramers doublets and spin-orbital coupling, the $4f$ electrons can be treated as $S_{eff}=1/2$ at low temperatures, and the spin-orbit entanglement can induce direction-dependent exchanges, such as the $K$ and $\Gamma$ terms. A typical example is YbMgGaO$_{4}$~\cite{srep16419,PhysRevLett.115.167203} and YbZnGaO$_{4}$~\cite{PhysRevLett.120.087201}, in which the Yb$^{3+}$ ions form a triangular layer and are surrounded by O$^{2-}$ which construct edge sharing octahedra. More recently, an alternative family of compounds AReCh$_2$ (A=alkali, Re=rare-earth, Ch=O, S, Se) with perfect triangular lattices of rare-earth ions have been synthesized and explored. The magnetic susceptibilty and heat capacity data suggest no long-range magnetic order or spin freezing down to the lowest measurement temperature, which implies their candidacy for QSL state~\cite{acsmaterialslett.9b00464,PhysRevMaterials.3.114413,PhysRevB.100.241116,Liu_2018,PhysRevB.99.180401,PhysRevB.100.224417,PhysRevB.100.144432}. These triangluar magnets of rare-earth ions provide suitable platforms to study the interplay of geometric frustation and exchange frustration induced by spin-orbit coupling.

The phase diagram of the HK model ($\Gamma=0$) on the triangular lattice has been studied by means of Luttinger-Tisza minimization together with classical Monte Carlo simulation~\cite{PhysRevB.93.104417}, exact diagonalization (ED)~\cite{PhysRevB.91.155135,KaiLi2015}, and density matrix renormalization group~\cite{JPSJ.85.114710,PhysRevX.9.021017}. All these methods give consistent results about four magnetically ordered phases: two collinear patterns of FM and stripe types, and two noncollinear patterns of 120$^\circ$ N\'{e}el and noncoplanar spiral types. Note that the distortions of the 120$^\circ$ N\'{e}el order, when the model deviates from the AFM Heisenberg limit ($K=0$, $J>0$), were also called the $Z_{2}$ vortex crystal~\cite{PhysRevB.93.104417,PhysRevB.91.155135,JPSJ.85.114710}. However, the nature of the phase around the antiferromagnetic Kitaev point is still under debate. The Luttinger-Tisza minimization method suggested it to be a $Z_{2}$ vortex crystal~\cite{PhysRevB.93.104417}, the Schwinger-fermion mean-field theory proposed it to be a QSL~\cite{KaiLi2015}, while the density matrix renormalization group calculations suggested it to be a nematic phase~\cite{PhysRevB.91.155135,JPSJ.85.114710} or a stripe phase~\cite{PhysRevX.9.021017}. When the symmetric off-diagonal $\Gamma$ interaction is included, a classical analysis reveals that the stripe and ferromagnetic phases dominate the $J$-$K$-$\Gamma$ phase diagram, in addition to small regions of 120$^\circ$ N\'{e}el, $Z_{2}$ vortex crystal and nematic phases~\cite{PhysRevB.92.165108}. However, the studies on the effects of quantum fluctuations on the global $J$-$K$-$\Gamma$ phase diagram are scarce. In particular, since no exact solution has been reported so far for the pure spin-$1/2$ Kitaev and $\Gamma$ models on the triangular lattice, it also remains conceptually interesting to investigate whether QSL states could exist as possible GSs due to quantum fluctuations introduced by these exchange-frustrated interactions.

In this paper, we study the global phase diagram and the phase transitions of the triangular lattice $J$-$K$-$\Gamma$ model using a combination of ED and classical analyses. We find that the ED results are basically consistent with the classical analyses such as the classical Monte Carlo simulations. In the HK limit ($\Gamma=0$), there are five classical phases: one FM, two stripe, one $120^{\circ}$ N\'{e}el and its dual phases. For the pure AFM Kitaev model, although the system has highly degenerate classical GSs composed of the stripe and nematic states, the order-by-disorder mechanism caused by quantum fluctuations makes the system select the stripe state to be the GS. When the $\Gamma$ term is included, we find five new phases: two FM, one stripe, one modulated stripe and one possible QSL phases. With the aid of classical analysis, we determine the spin directions for each magnetically ordered phase and explain that why there are phase transitions between the phases with the same type of classical orders. On the other hand, we find that the order-by-disorder mechanism makes the GSs of the pure $\Gamma$ models have FM orders, although the classical analyses suggest that the GSs are highly degenerate. For the possible QSL phase, based on the investigation of the spin excitation spectrum, we find that the GS is probably a gapped $Z_{2}$ QSL.

@@

In the followings, for convenience, we fix the energy scale with $\sqrt{J^2 + K^2 + \Gamma^2}=1$ and parametrize the exchange parameters using spherical angles $\alpha$ and $\beta$
\begin{equation}
    J = \sin\alpha \sin\beta, \quad
    K = \sin\alpha \cos\beta, \quad
    \Gamma = \cos\alpha
    \label{eq:Parameters}
\end{equation}
where $\alpha \in [0, \pi]$ and $\beta \in [0, 2\pi]$ to cover the global parameter space.

In the HK limit ($\Gamma=0$), the model~\eqref{eq:Hamiltonian} admits an exact duality transformation, i.e., the so-called four-sublattice-transformation (FST)~\cite{PhysRevB.89.014414}. The FST is a spin rotation transformation which divides the triangluar lattice into four sublattices (see Fig.~\ref{fig:ModelDefinition}(c)) and performs the following rotations of the spins on the four sublattices to map the spin $\bm{S}_{i}$ to $\bm{S}_{i}^{\prime}$,
\begin{align*}
    & \bm{S}_{0}^{\prime} = \bm{S}_{0}& \quad \text{for sublattice 0}, \\
    & \bm{S}_{1}^{\prime} = (S_1^x, -S_1^y, -S_1^z)& \quad \text{for sublattice 1}, \\
    & \bm{S}_{2}^{\prime} = (-S_2^x, S_2^y, -S_2^z)& \quad \text{for sublattice 2}, \\
    & \bm{S}_{3}^{\prime} = (-S_3^x, -S_3^y, S_3^z)& \quad \text{for sublattice 3}.
\end{align*}
This corresponds to a $\pi$ rotation around $x$, $y$, and $z$ axis for the spin operators on the sublattices $1$, $2$, and $3$, respectively. The resulting Hamiltonian $H^{\prime}(\bm{S}^{\prime})$ has the same form as the original Hamiltonian albeit with different model parameters $J^{\prime} = -J$ and $K^{\prime} = 2J + K$. For the spherical angles defined in Eq.~\eqref{eq:Parameters}, the mapping takes the form
\begin{align}
    \tan\beta^{\prime} = -\sin\beta / (2\sin\beta + \cos\beta).
    \label{eq:FST}
\end{align}
This special property of the model can help us to identify some exotic magnetically ordered phases from the well established simple counterparts.

@@

For the other two phases with large areas and mainly located in the $J<0$ region of the phase diagram, we show the SSFs of two representative points with $\alpha=0.3\pi, \beta=1.3\pi$ and $\alpha=0.7\pi, \beta=1.65\pi$ in Fig.~\ref{fig:StructureFactors}(f) and (g), respectively. Since the SSFs are peaked at $\tilde{\Gamma}$ point of the first BZ, so both phases have ferromagnetic orders and they are denoted with FM-A and FM-C in Fig.~\ref{fig:QuantumPhaseDiagram}. We note that there are two special points with $\Gamma=\pm1$ ($J=K=0$) in the two FM phases, which are located at the phase boundary in the classical phase diagram~\cite{PhysRevB.92.165108}. Thus, it is worthy to discuss the properties of the pure $\Gamma$ models in details, and we perform the numerical simulations by using the ``basin-hopping" global optimization and classical Monte Carlo methods. For $\Gamma=+1$, apart from the FM GS with the spin orientation lying in the lattice plane, we also found disordered states having the same energy as the FM state. For $\Gamma=-1$, in addition to the FM order perpendicular to the lattice plane, other long-range ordered states which are energetically degenerate with the FM state also appear (see Appendix~\ref{apx:DegeneratedStates} for details). However, for the quantum model, quantum flucations select the FM ordered states as the GS out of the degenerate manifolds of classical states according to our ED calculations. This selection of states among the degenerate classical ground states is the so the called order-by-disorder mechanism, which has previously been applied in a number of insulating magnets, especially in cases where frustration leads to a degenerate manifold of classical ground state configurations that is broken by quantum fluctuations~\cite{JPSJ.54.4494,PhysRevLett.62.2056,PhysRevLett.88.067203,PhysRevB.81.214419,PhysRevLett.109.077204}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{fig/StructureFactors.pdf}
    \caption{\label{fig:StructureFactors}(Color online) (a)-(h) Representative SSFs from the ED calculations for different interaction parameters. The green dashed lines marks the first BZ corresponding to the triangular lattice. (i) Characteristic wave vectors for the FM and stipe phases. See Appendix~\ref{apx:SSF} for the SSFs of the 120$^\circ$ N\'{e}el phase, dual N\'{e}el phase and pure $\Gamma$ models.}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{fig/Probabilities.pdf}
    \caption{\label{fig:Proabilities}(Color online) Map of the probabilities of the cluster spin coherent state given by Eq.~\eqref{eq:ClusterCoherentState} in exact cluster GS for (a) FM-A, (b) FM-C, (c) stripe-A and (d) stripe-B phases. The radial and polar coordinate gives the angles $\theta$ and $\phi$, which are spherical angles with respect to the reference frame shown in Fig.~\ref{fig:ModelDefinition}(b). The green dashed lines and solid circles mark the ordered moment directions for the classical states. Note that the GSs are six-fold degenerate for the stripe-A and stripe-B phase, and two-fold degenerate for the FM-C phase, so the total probabilities for all of these phases are approximate to $1$. }
\end{figure}

>>>>

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{fig/ModelDefinition.pdf}
    \caption{\label{fig:ModelDefinition}(Color online) (a) Top view of the triangular lattice of the edge-sharing octahedron. (b) The orientation of the cubic $x$, $y$, $z$ axes with respect to the octahedron. The spin operators $S^x$, $S^y$ and $S^z$ are defined with respect to this reference frame. (c) Three types of the NN bonds on the triangular lattice, namely $\gamma=x, y, z$ colored red, green and blue, respectively. The different colors of the lattice sites label the four sublattices realizing the four sublattice transformation ( see main text for details).}
\end{figure}
In fact, magnetic ions located at the center of edge-shared octahedra can not only form the honeycomb lattice but also the triangular lattice (see Fig.~\ref{fig:ModelDefinition}(a)), so the Kitaev and $\Gamma$ terms can naturally be generalized to the triangular lattice~\cite{PhysRevB.89.014414,PhysRevB.93.104417}.
On the experimental side, studies of several classes of rare-earth compounds have recently shown that the quantum spin model on the triangular lattice can be formed by $4f$ moments. In particular, due to their possible Kramers doublets and spin-orbital coupling, the $4f$ electrons can be treated as $S_{eff}=1/2$ at low temperatures, and the spin-orbit entanglement can induce direction-dependent exchanges, such as the $K$ and $\Gamma$ terms. A typical example is YbMgGaO$_{4}$~\cite{srep16419,PhysRevLett.115.167203} and YbZnGaO$_{4}$~\cite{PhysRevLett.120.087201}, in which the Yb$^{3+}$ ions form a triangular layer and are surrounded by O$^{2-}$ which construct edge sharing octahedra. More recently, an alternative family of compounds AReCh$_2$ (A=alkali, Re=rare-earth, Ch=O, S, Se) with perfect triangular lattices of rare-earth ions have been synthesized and explored. The magnetic susceptibilty and heat capacity data suggest no long-range magnetic order or spin freezing down to the lowest measurement temperature, which implies their candidacy for QSL state~\cite{acsmaterialslett.9b00464,PhysRevMaterials.3.114413,PhysRevB.100.241116,Liu_2018,PhysRevB.99.180401,PhysRevB.100.224417,PhysRevB.100.144432}. These triangluar magnets of rare-earth ions provide suitable platforms to study the interplay of geometric frustation and exchange frustration induced by spin-orbit coupling.

The phase diagram of the HK model ($\Gamma=0$) on the triangular lattice has been studied by means of Luttinger-Tisza minimization together with classical Monte Carlo simulation~\cite{PhysRevB.93.104417}, exact diagonalization (ED)~\cite{PhysRevB.91.155135,KaiLi2015}, and density matrix renormalization group~\cite{JPSJ.85.114710,PhysRevX.9.021017}. All these methods give consistent results about four magnetically ordered phases: two collinear patterns of FM and stripe types, and two noncollinear patterns of 120$^\circ$ N\'{e}el and noncoplanar spiral types. Note that the distortions of the 120$^\circ$ N\'{e}el order, when the model deviates from the AFM Heisenberg limit ($K=0$, $J>0$), were also called the $Z_{2}$ vortex crystal~\cite{PhysRevB.93.104417,PhysRevB.91.155135,JPSJ.85.114710}. However, the nature of the phase around the antiferromagnetic Kitaev point is still under debate. The Luttinger-Tisza minimization method suggested it to be a $Z_{2}$ vortex crystal~\cite{PhysRevB.93.104417}, the Schwinger-fermion mean-field theory proposed it to be a QSL~\cite{KaiLi2015}, while the density matrix renormalization group calculations suggested it to be a nematic phase~\cite{PhysRevB.91.155135,JPSJ.85.114710} or a stripe phase~\cite{PhysRevX.9.021017}. When the symmetric off-diagonal exchange interaction $\Gamma$ is included, a classical analysis reveals that the stripe and ferromagnetic phases dominate the $J$-$K$-$\Gamma$ phase diagram, in addition to small regions of 120$^\circ$, $Z_{2}$ vortex crystal and nematic phases~\cite{PhysRevB.92.165108}. However, the studies on the effects of quantum fluctuations on the global $J$-$K$-$\Gamma$ phase diagram are scarce. In particular, since no exact solution has been reported so far for the pure spin-$1/2$ Kitaev and $\Gamma$ models on the triangular lattice, it also remains conceptually interesting to investigate whether QSL states could exist as possible GSs due to quantum fluctuations introduced by these exchange-frustrated interactions.

In this paper, we study the global phase diagram and the phase transitions of the triangular-lattice $J$-$K$-$\Gamma$ model using a combination of ED and classical analyses. We find that the ED results are basically consistent with the classical analyses such as the classical Monte Carlo simulations. In the HK limit ($\Gamma=0$), there are five classical phases: one FM, two stripe, one $120^{\circ}$ N\'{e}el and its dual phases. For the pure AFM Kitaev model, although the system has highly degenerate classical GSs composed of the stripe and nematic states, the order-by-disorder mechanism caused by quantum fluctuations makes the system select the stripe state to be the GS. When the $\Gamma$ term is included, we find five new phases: two FM, one stripe, one modulated stripe and one possible QSL phases. With the aid of classical analysis, we determine the spin directions for each magnetically ordered phase and explain that why there are phase transitions between the phases with the same type of classical orders. On the other hand, we find that the order-by-disorder mechanism makes the GSs of the pure $\Gamma$ models have FM orders, although the classical analyses suggest that the GSs are highly degenerate. For the possible QSL phase, based on the investigation of the spin excitation spectrum, we find that the GS is probably a gapped $Z_{2}$ QSL.

@@

In the followings, for convenience, we fix the energy scale with $\sqrt{J^2 + K^2 + \Gamma^2}=1$ and parametrize the exchange interactions using spherical angles $\alpha$ and $\beta$
\begin{equation}
    J = \sin\alpha \sin\beta, \quad
    K = \sin\alpha \cos\beta, \quad
    \Gamma = \cos\alpha
    \label{eq:Parameters}
\end{equation}
where $\alpha \in [0, \pi]$ and $\beta \in [0, 2\pi]$ to cover the global parameter space.

In the HK limit ($\Gamma=0$), the model~\eqref{eq:Hamiltonian} admits an exact duality transformation, i.e., the so-called four-sublattice-transformation (FST)~\cite{PhysRevB.89.014414}. The FST is a spin rotation transformation which divides the triangluar lattice into four sublattices (see Fig.~\ref{fig:ModelDefinition}(c)) and performs the following rotations of the spins on the four sublattices to map the spin $\bm{S}_{i}$ to $\bm{S}_{i}^{\prime}$,
\begin{align*}
    & \bm{S}_{0}^{\prime} = \bm{S}_{0}& \quad \text{for sublattice 0}, \\
    & \bm{S}_{1}^{\prime} = (S_1^x, -S_1^y, -S_1^z)& \quad \text{for sublattice 1}, \\
    & \bm{S}_{2}^{\prime} = (-S_2^x, S_2^y, -S_2^z)& \quad \text{for sublattice 2}, \\
    & \bm{S}_{3}^{\prime} = (-S_3^x, -S_3^y, S_3^z)& \quad \text{for sublattice 3}.
\end{align*}
This corresponds to a $\pi$ rotation around $x$, $y$, and $z$ axis for the sublattices $1$, $2$, and $3$, respectively. The resulting Hamiltonian $H^{\prime}(\bm{S}^{\prime})$ has the same form as the original Hamiltonian albeit with different model parameters $J^{\prime} = -J$ and $K^{\prime} = 2J + K$. For the spherical angles defined in Eq.~\eqref{eq:Parameters}, the mapping takes the form
\begin{align}
    \tan\beta^{\prime} = -\sin\beta / (2\sin\beta + \cos\beta).
    \label{eq:FST}
\end{align}
This special property of the model can help us to identify some exotic magnetically ordered phases from the well established simple counterparts.

@@

For the other two phases with large areas and mainly located in the $J<0$ region of the phase diagram, we show the SSFs of two representative points with $\alpha=0.3\pi, \beta=1.3\pi$ and $\alpha=0.7\pi, \beta=1.65\pi$ in Fig.~\ref{fig:StructureFactors}(f) and (g), respectively. Since the SSFs are peaked at $\tilde{\Gamma}$ point of the first BZ, so both phases have ferromagnetic orders and they are denoted with FM-A and FM-C in Fig.~\ref{fig:QuantumPhaseDiagram}. We note that there are two special points with $J=K=0$ and $\Gamma=\pm1$ in the two FM phases, which are located at the phase boundary in the classical phase diagram~\cite{PhysRevB.92.165108}. Thus, it is worthy to discuss the properties of the pure $\Gamma$ models in details, and we perform the numerical simulations by using the ``basin-hopping" global optimization and classical Monte Carlo methods. For $\Gamma=+1$, apart from the FM GS with the spin orientation lying in the lattice plane, we also found disordered states having the same energy as the FM state. For $\Gamma=-1$, in addition to the FM order perpendicular to the lattice plane, other long-range ordered states which are energetically degenerate with the FM state also appear (see Appendix~\ref{apx:DegeneratedStates} for details). However, for the quantum model, quantum flucations select the FM ordered states as the GS out of the degenerate manifolds of classical states according to our ED calculations. This selection of states among the degenerate classical ground states is the so the called order-by-disorder mechanism, which has previously been applied in a number of insulating magnets, especially in cases where frustration leads to a degenerate manifold of classical ground state configurations that is broken by quantum fluctuations~\cite{JPSJ.54.4494,PhysRevLett.62.2056,PhysRevLett.88.067203,PhysRevB.81.214419,PhysRevLett.109.077204}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{fig/StructureFactors.pdf}
    \caption{\label{fig:StructureFactors}(Color online) (a)-(h) Representative SSFs from the ED calculations for different interaction parameters. The green dashed lines marks the first BZ corresponding to the triangular lattice. (i) Characteristic wave vectors for the FM and stipe phases. See Appendix~\ref{apx:SSF} for the SSFs of the 120$^\circ$ N\'{e}el phase, dual N\'{e}el phase and pure $\Gamma$ models.}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{fig/Probabilities.pdf}
    \caption{\label{fig:Proabilities}(Color online) Map of the probabilities of the cluster spin coherent state given by Eq.~\eqref{eq:ClusterCoherentState} in exact cluster GS for (a) FM-A, (b) FM-B, (c) stripe-A and (d) stripe-B phases. The radial and polar coordinate gives the angles $\theta$ and $\phi$, which are spherical angles with respect to the reference frame shown in Fig.~\ref{fig:ModelDefinition}(b). The green dashed lines and solid circles mark the ordered moment directions for the classical states. Note that the GSs are six-fold degenerate for the FM-A, stripe-A and stripe-B phase, and two-fold degenerate for the FM-B phase, so the total probabilities for all of these phases are approximate to $1$. }
\end{figure}
